---
title: "Reproducible Research: Peer Assessment 1"
author: "Herbert Amaya"
output: 
  html_document: 
    keep_md: yes
keep_md: true
---


## Loading and preprocessing the data
```{r global_options, echo=TRUE}
knitr::opts_chunk$set(fig.path='figure/',echo=TRUE)
data <- read.csv(unz("activity.zip","activity.csv"))
library(dplyr)
```


## What is mean total number of steps taken per day?
```{r, echo=TRUE}
data.dailyStepsAvg <- group_by(data,date)
data.dailyStepsAvg <- summarize(data.dailyStepsAvg,steps=sum(steps,na.rm=TRUE))
hist(data.dailyStepsAvg$steps,main = "Histogram of Total Number of Steps taken each day",xlab = "Steps per Day")
```

### Mean and Median Calculation
```{r, echo=TRUE}
mean(data.dailyStepsAvg$steps,na.rm=TRUE)
median(data.dailyStepsAvg$steps,na.rm = TRUE)
```

## What is the average daily activity pattern?
```{r, echo=TRUE}
data.intervalAvg <- group_by(data,interval)
data.intervalAvg <- summarize(data.intervalAvg,avg_steps=mean(steps,na.rm=TRUE))
library(ggplot2)
ggplot(data.intervalAvg,aes(interval,avg_steps)) + geom_line()

```

### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r, echo=TRUE}
fiveMin.Interval.MaxSteps = arrange(data.intervalAvg,desc(avg_steps))[1,]
print.data.frame(fiveMin.Interval.MaxSteps)
```

## Imputing missing values
```{r, echo=TRUE}
missingValues <- sapply(data, function(x) sum(is.na(x)))
names(missingValues)[1]="Total missing values"
print(missingValues[1])
```
Filling in the missing values with the mean for the 5-minute interval, and creating a new data set.
```{r, echo=TRUE}
data.noMissing <- 
    mutate(data,
           steps=ifelse(is.na(steps),
                        data.intervalAvg[which(data.intervalAvg$interval == interval),]$avg_steps
                        ,steps))
```
Histogram of the total number of steps taken each day
```{r, echo=TRUE}
data.dailyStepsAvg <- group_by(data.noMissing,date)
data.dailyStepsAvg <- summarize(data.dailyStepsAvg,steps=sum(steps,na.rm=TRUE))
hist(data.dailyStepsAvg$steps,main = "Histogram of Total Number of Steps taken each day",xlab = "Steps per Day")

```

Mean and Median total number of steps taken per day
```{r, echo=TRUE}
mean(data.dailyStepsAvg$steps,na.rm=TRUE)
median(data.dailyStepsAvg$steps,na.rm = TRUE)
```

###Do these values differ from the estimates from the first part of the assignment? 
Yes, they slightly differ.

###What is the impact of imputing missing data on the estimates of the total daily number of steps?
Imputing missing data increases the amount of 'valid data' for the analysis, thus slightly weighting on the final results.

## Are there differences in activity patterns between weekdays and weekends?
```{r, echo=TRUE}
weekend.labels <- c("Saturday","Sunday")
day_type.levels <- c("weekend","weekday")
data.dayType <- mutate(data.noMissing,day_type = factor(ifelse(weekdays(as.Date(date)) %in% weekend.labels,"weekend","weekday"),levels = day_type.levels))

data.dayType.intervalAvg <- group_by(data.dayType,interval,day_type)
data.dayType.intervalAvg <- summarize(data.dayType.intervalAvg,avg_steps=mean(steps,na.rm=TRUE))

q <- qplot(interval,avg_steps,data=data.dayType.intervalAvg,facets=.~day_type,geom ="line") + facet_wrap( ~ day_type, nrow=2) 
print(q)
```

During weekends the distribution of steps across the day seems to be more uniform than weekdays, for which there is a peak of steps from 750 to 1000 interval, follow by a sustantial decrease of steps for the rest of the day, with a small uptick from 1750 to 1900 interval.